Composition-climate paper

### 2016-08-23 Composition_Ecology.Rmd
Revisions up through the Results now.  Only the Discussion & references to work through.  Some figure work too, likely.

Regionally, $T_{min}$ has increased by `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal `r fig_nums("land_use_clim", display = "cite")`, $T_{max}$ has declined (`r round(clim_change[2], 1)`^o^C).  Because of the large shift in minimum temperatures, $T_{diff}$ decreased by `r round(clim_change[3], 1)`^o^C.  Annual precipitation in the region has increased slightly (`r round(clim_change[1], 1)`mm).  The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicate no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.

@@

```{r echo = FALSE}
tab_nums("tab_gain_loss", "Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns are percentages relative to the total area occupied by the taxon in the presettlement and modern eras. Modern and Historical indicate represent range area (100s of km^2^) occupied by the tree taxon for the pre-settlement or modern era.  Taxa are sorted by the percent area lost (Column 3).", display = FALSE)
```

For most tree genera, historic range losses dominate over range gains and are spatially structured (Figure 3). Range losses in the Upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For *Ulmus* and *Larix*, however, losses occur throughout the range.  Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.   Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Thuja*, and *Ulmus* gains are limited but primarily in the southern portion of their historic range, or beyond it.

@@

Changes in spatial coverage and changes in regional climate result in overall changes in the climate envelopes occupied by genera in the upper Midwest (`r fig_nums("boxplots", display = "cite")`).

@@

The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* in Figure 2.  For *Larix* we see clear shifts in climate space (from solid to dashed lines) whether modern or presettlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.  Historical land use also plays a role in shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern, however, land use appears to play a smaller role for the other climate variables.

Shifts in tree-climate distributions are significant for all genera and all climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons (all p < 0.0001).  The directionality of shifts between historical and modern climate distributions are consistent across genera: modern $T_{max}$ is lower for all genera, while $T_{min}$ is higher (`r fig_nums("boxplot", display = "cite")`).  Across all genera modern annual precipitation is also higher than in the past.  *Larix* and *Ulmus* show the most extreme changes in climate space overall.  *Ulmus* shows the largest increase in minimum temperature ($\delta_{Tmin}$ = 4.2 ^o^C) and the smallest increase occurs for *Larix*. Maximum annual temperature declines most for *Ostrya/Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).  *Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), the smallest change is for *Larix* ($\delta_{Pann}$ = 42mm).

@@

Both the total change in realized climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5). Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).   Attribution analyses based on the $d$ index (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa. The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus*. The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historic climate change. For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for other taxa are driven by changing climate.  The climatic signal is strongest for $P_{ann}$, with regional changes being largely responsible for shifting climate niches (`r fig_nums("hellPlot", display = "cite")`), while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than the climate shifts.

@@

The majority of climate - land-use interactions are compounding (n = 30) or confounding (n = 28), while few (n = 2) show evidence of counteracting effects (`r tab_nums("compTable", display = "cite")`).  Compounding interactions are particularly strong for $T_{max}$ (Table 3), and are consistent with the observed trend of lowered $T_{min}$ (Figure 1) and losses concentrated in the souther portions of species ranges (Figure 2); both tend to shift tree-climate distributions to cooler portions of climate space. These analyses are insenstitive to the type of correction applied to the *p*-value, although running the analysis without a correction factor reduces the counts of confounding relationships and increases counts in both counteracting and compounding interactions.  Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\Chi^{2}}_{4}$ = 15.22), but are not significantly correlated for other climate variables.

For climate variables, the proportion of compounding interactions is greatest for $T_{max}$, and least for $T_{min}$, for which confounding interactions dominate.  For $P_{ann}$ and $T_{diff}$ we see a general balance of compounding and confounding interactions.  Only *Ulmus* (for $T_{max}$) and *Larix* (for $T_{min}$) show counteracting changes for climate/land-use interactions (`r tab_nums("compTable", display = "cite")`).

Several genera show common patterns in their relationships to climate/land-use interactions.  *Populus*, *Picea*, *Quercus*, *Acer* and *Betula* all show compounding relationships for both $P_{ann}$ and $T_{max}$ and either compounding or confounding interactions for $T_{diff}$.  All five taxa show confounding interactions for $T_{min}$ (`r tab_nums("compTable", display = "cite")`).  These taxa all show significant losses in range along the southern and western margins of their distribution (`r fig_nums("gain_loss", display="cite")`), with little new distribution in those regions.

*Fraxinus*, *Tsuga* and *Tilia* all show confounding interactions with $P_{ann}$, $T_{diff}$ and $T_{min}$, with compounding interactions with $T_{max}$.  Each of these taxa show losses throughout historical distribution, but without clear spatial structure (`r fig_nums("gain_loss", display="cite")`). No taxa show patterns that are identical to another genera.

>>>>

Regionally, $T_{Jan}$ increased `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal (`r fig_nums("land_use_clim", display = "cite")`), $T_{Jul}$ declined `r round(clim_change[2], 1)`^o^C.  Because of the large shift in minimum temperatures, $T_{diff}$ decreased by `r round(clim_change[3], 1)`^o^C. Regionally, $P_{ann}$ has increased by `r round(clim_change[1], 1)`mm.  The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.

@@

```{r echo = FALSE}
tab_nums("tab_gain_loss", "Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns are percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras. Modern and Historical indicate represent range area (100s of km^2^) occupied by the tree taxon for the pre-settlement or modern era.  Taxa are sorted by the percent area lost (Column 3).", display = FALSE)
```

For most tree genera, historic range losses dominate over range gains and are spatially structured (Table 1, Figure 3). Range losses in the upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For *Ulmus* and *Larix*, however, losses occur throughout the range.  Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.   Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historic range, or beyond it.

@@

@@

The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* in Figure 2.  For *Larix* we see clear shifts in climate space when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.  Historical land use also plays a role in shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern (Fig. 2d). However, for *Larix*, land use overall appears to play a smaller role than climate change.

Shifts in tree-climate distributions are significant for all genera and all climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons (all p < 0.0001).  The directionality of shifts between historical and modern climate distributions is consistent across genera: modern $T_{max}$ is lower for all genera, while $T_{min}$ is higher (`r fig_nums("boxplot", display = "cite")`).  Across all genera, modern $P_{ann}$ is higher than in the past. *Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{Tmin}$ = 4.2 ^o^C), while the smallest increase occurs for *Larix*. Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).  *Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), while the smallest change is for *Larix* ($\delta_{Pann}$ = 42mm).

@@

Both the total shift in realized tree-climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5). Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).   Attribution analyses based on $d$ (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa. The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches (), while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than climate shifts. The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus*. The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historic climate change. For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributable to changing climate.

@@

The majority of climate - land-use interactions are compounding (n = 30) or confounding (n = 28), while few (n = 2) show counteracting effects (`r tab_nums("compTable", display = "cite")`).  Compounding interactions are particularly strong for $T_{max}$ (Table 3), and are consistent with the observed trend of lowered $T_{min}$ (Figure 1) and losses concentrated in the southern portions of species ranges (Figure 2); both tend to shift tree-climate distributions to cooler portions of climate space. Compounding interactions are also strong for $P_{ann}$ and are caused by both land use and climate change shifting species to a wetter portion of climate space than they occupied historically.  These analyses are insensitive to the type of correction applied to the *p*-value, although running the analysis without a correction factor reduces the counts of insignificant (confounding) relationships and increases the number of  counteracting and compounding interactions.  Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\chi^{2}}_{4}$ = 15.22), but are not significantly correlated for other climate variables.

Several genera show common patterns in their relationships to climate/land-use interactions.  *Populus*, *Picea*, *Quercus*, *Acer* and *Betula* all show compounding relationships for both $P_{ann}$ and $T_{max}$, either compounding or confounding interactions for $T_{diff}$. and confounding interactions for $T_{min}$ (`r tab_nums("compTable", display = "cite")`).  These taxa all show significant losses in range along the southern and western margins of their distribution (`r fig_nums("gain_loss", display="cite")`), with few gains in those regions.

*Fraxinus*, *Tsuga* and *Tilia* all show confounding interactions with $P_{ann}$, $T_{diff}$ and $T_{min}$, with compounding interactions with $T_{max}$.  Each of these taxa show losses throughout historical distribution, but without clear spatial structure (`r fig_nums("gain_loss", display="cite")`). No taxa show patterns that are identical to another genera. Only *Ulmus* (for $T_{max}$) and *Larix* (for $T_{min}$) show counteracting changes for climate/land-use interactions (`r tab_nums("compTable", display = "cite")`).
